---
Title: "Data Analysis"
Author: Hunter Everton
---

## Analyzing the dataset

Load the necessary libraries
```{r load_libraries, message = FALSE}
library(here)
library(tidyverse)
library(ggplot2)
library(sf)
library(terra)
```

Load the dataset

```{r, message = FALSE}
oba_land_use <- readRDS(file = here("data", "cleaned_data", "oba_land_use.rds"))

head(oba_land_use)
```

We can see that the final combined dataset has columns for the longitude and latitude of each bee record, each diversity index and the proportion of 
each land type. For making analysis easier I will combine the undeveloped land types into one measure of undeveloped land and the developed land intensities
into one developed land proportion.

```{r}
oba_land_use_mutated <- oba_land_use %>%
    mutate(prop_undeveloped = prop_open_water +
                              prop_perennial_ice_snow +
                              prop_barren_land +
                              prop_deciduous +
                              prop_evergreen +
                              prop_mixed_forest +
                              prop_shrub_scrub +
                              prop_grassland_herbaceous +
                              prop_pasture_hay +
                              prop_cultivated_crops +
                              prop_woody_wetlands,
           prop_developed = prop_developed_open_space +
                            prop_developed_low_intensity +
                            prop_developed_med_intensity +
                            prop_developed_high_intensity) %>%
    select(place_name, num_unique, shannon, simpson, inv_simpson,
           prop_developed, prop_undeveloped,
           longitude, latitude)
```

## Summary Statistics

```{r}
ggplot(oba_land_use_mutated) +
    geom_histogram(aes(x = shannon), color = "white", fill = "black")

ggplot(oba_land_use_mutated) +
    geom_histogram(aes(x = simpson), color = "white", fill = "black")

ggplot(oba_land_use_mutated) +
    geom_histogram(aes(x = inv_simpson), color = "white", fill = "black")

ggplot(oba_land_use_mutated) +
    geom_histogram(aes(x = prop_developed),
                   color = "white", fill = "black") +
    labs(x = "proportion developed land")
```

## Regression Analysis

```{r}
shannon_plot <- ggplot(oba_land_use_mutated, aes(x = prop_developed, y = shannon)) +
    geom_point() +
    geom_smooth(method = "lm") +
    labs(title = "Shannon Diversity", x = "proportion developed land", y = "shannon diversity index") +
    theme_bw() +
    theme(axis.text = element_text(size = 17),
          text = element_text(size = 17))

ggsave(here("analysis", "figures", "shannon.png"), shannon_plot, width = 8, height = 8)

simpson_plot <- ggplot(oba_land_use_mutated, aes(x = prop_developed, y = simpson)) +
    geom_point() +
    geom_smooth(method = "lm") +
    labs(title = "Simpson Diversity", x = "proportion developed land", y = "simpson diversity index") +
    theme_bw() +
    theme(axis.text = element_text(size = 17),
          text = element_text(size = 17))

ggsave(here("analysis", "figures", "simpson.png"), simpson_plot, width = 8, height = 8)
```

```{r}
m_shannon <- lm(shannon ~ prop_developed, data = oba_land_use_mutated)
summary(m_shannon)

m_simpson <- lm(simpson ~ prop_developed, data = oba_land_use_mutated)
summary(m_simpson)
```

## Spatial Maps

```{r}
or <- map_data("state", "oregon") %>%
    select(lon = long, lat, group, id = subregion)

oba_land_use_spatial <- oba_land_use_mutated %>%
    unnest(cols = c(longitude, latitude)) %>%
    st_as_sf(coords = c("longitude", "latitude"),
             crs = crs("EPSG:4326"))

developed_spatial <- ggplot() +
    geom_polygon(data = or, aes(lon, lat, group = group), fill = "white", color = "black") +
    geom_sf(data = oba_land_use_spatial, aes(color = prop_developed)) +
    scale_color_viridis_c() +
    labs(x = "", y = "", color = "proportion developed land") +
    theme_void() +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.title = element_text(size = 17),
          legend.text = element_text(size = 11, angle = -45),
          legend.box.margin = margin(-10, -10, -10, -10),
          legend.position = "bottom")

ggsave(here("analysis", "figures", "developed_spatial.png"), developed_spatial)

num_unique_spatial <- ggplot() +
    geom_polygon(data = or, aes(lon, lat, group = group), fill = "white", color = "black") +
    geom_sf(data = oba_land_use_spatial, aes(color = num_unique)) +
    scale_color_viridis_c() +
    labs(x = "", y = "", color = "number of unique species") +
    theme_void() +
    theme(axis.text.x = element_blank(),
          axis.ticks.x = element_blank(),
          axis.text.y = element_blank(),
          axis.ticks.y = element_blank(),
          legend.title = element_text(size = 17),
          legend.text = element_text(size = 11),
          legend.box.margin = margin(-10, -10, -10, -10),
          legend.position = "bottom")

ggsave(here("analysis", "figures", "num_unique_spatial.png"), num_unique_spatial)
```