---
Title: "Cleaning Data"
Author: Hunter Everton
---

# Cleaning Data

Load the necessary libraries
```{r load_libraries, message = FALSE}
library(here)
library(tidyverse)
library(stringr)
library(vegan)
```

Load the datasets
```{r}
oba_data <- read_csv(here("data", "raw_data", "OBA_2018-2024.csv"))
land_use <- read_csv(here("data", "raw_data", "nhgis_place2022_tl2022_nlcd2021.csv"))
or_places <- read_delim(here("data", "raw_data", "or_places.txt"), delim = "|")

head(oba_data)
head(land_use)
head(or_places)
```

## Cleaning Land Use Data

First I'll work on the land use data as that requires more cleaning. The land use dataset contains all cities and incorporated places in the U.S.,
so first let's subset the data to just those from Oregon. The dataset from NHGIS uses FIPS codes in its GISJOIN variable; using the table of FIPS
codes for Oregon, which can be found [here](https://www2.census.gov/geo/docs/reference/codes2020/place/st41_or_place2020.txt), I will subset the
GISJOIN variable to the Oregon FIPS codes.

```{r}
# Find the row where the Oregon places start
or_start <- which(land_use$GISJOIN == "G41000275")
# Find the end of the Oregon places
or_end <- which(land_use$GISJOIN == "G41084600")

# Subset the land use data to the Oregon rows
land_use <- land_use[or_start : or_end, ]

# Check that the land use data and place names have the same number of places
nrow(land_use) == nrow(or_places)
```

Now that I have the land use dataset subset to just the Oregon places I want to add the place names to the land use dataset. I then want to join the
datasets on the GISJOIN variable.

```{r}
# Add the GISJOIN column structure from the land use data to the Oregon places dataset
or_places$GISJOIN <- paste("G", or_places$STATEFP, "0", or_places$PLACEFP, sep = "")

# Make sure they are the same, this should be the same as the number of rows in either dataset
sum(or_places$GISJOIN == land_use$GISJOIN) == nrow(land_use)

# Join on GISJOIN variable and select columns needed
land_use <- land_use %>%
    left_join(or_places, by = "GISJOIN") %>%
    select("PLACENAME", "COUNTIES", starts_with("PROP"))

# Rename all columns
colnames(land_use) <- c("place_name",
                        "county",
                        "prop_open_water",
                        "prop_perennial_ice_snow",
                        "prop_developed_open_space",
                        "prop_developed_low_intensity",
                        "prop_developed_med_intensity",
                        "prop_developed_high_intensity",
                        "prop_barren_land",
                        "prop_deciduous",
                        "prop_evergreen",
                        "prop_mixed_forest",
                        "prop_shrub_scrub",
                        "prop_grassland_herbaceous",
                        "prop_pasture_hay",
                        "prop_cultivated_crops",
                        "prop_woody_wetlands",
                        "prop_emergent_herbaceous_wetlands")

head(land_use)
```

The place names end in either "city" or "CDP", and the places across multiple counties are denoted with the separator "~~~". These values should
be removed to merge with the oba data.

```{r}
land_use$place_name <- str_remove(land_use$place_name, "\\s(city|CDP)$")
land_use$county <- str_remove(land_use$county, "~~~.*")
land_use$county <- str_remove(land_use$county, "\\s(County)")
```

The land use data is now cleaned and ready for merging. Next I will clean the oba data.

## Cleaning OBA Data

```{r}
# Subset to Oregon data
oba_data_oregon <- oba_data[oba_data$stateProvince == "OR", ]

# Subset to places in the land use dataset
oba_data_places <- oba_data_oregon[oba_data_oregon$locality %in% land_use$place_name, ]

# Select needed columns
oba_data_subset <- oba_data_places %>%
    select(locality,
           genus,
           specificEpithet,
           decimalLatitude,
           decimalLongitude) %>%
    filter(!is.na(specificEpithet) & !is.na(locality)) %>%
    mutate(species = paste(genus, specificEpithet)) %>%
    select(-genus, -specificEpithet)

# Sort the data by locality in alphabetical order
oba_data_subset <- arrange(oba_data_subset, locality)

# Count the species for each locality
# Pivot the data to have each species in a column
oba_data_counts <- oba_data_subset %>%
    count(locality, species) %>%
    pivot_wider(names_from = species,
                values_from = n)

# Turn the data into a matrix for calculating diversity
oba_matrix <- as.matrix(oba_data_counts[, -1])

rownames(oba_matrix) <- unique(oba_data_subset$locality)

dim(oba_matrix)
```

Now I have a matrix where each row represents a location in both the place and OBA datasets and each column represents a species of bee in the OBA dataset.
Next I will compute diversity metrics of this matrix and add start creating the final dataset.

## Creating the final dataset

```{r}
# Get unique species for each locality
oba_data_unique <- oba_data_subset %>%
    group_by(locality) %>%
    summarize(num_unique = n_distinct(species))

# Change NA values to be 0
oba_matrix[is.na(oba_matrix)] <- 0

# Find the shannon diversity index for each locality
shannon <- diversity(oba_matrix, index = "shannon")

# Find the simpson diversity index
simpson <- diversity(oba_matrix, index = "simpson")

# Find the inverse simpson diversity index
inv_simpson <- diversity(oba_matrix, index = "invsimpson")
```

```{r}
# Set the columns of the final dataset to the diversity indices gotten earlier
# and the place names from the oba dataset
final_dataset_template <- data.frame(
    place_name = unique(oba_data_subset$locality),
    num_unique = oba_data_unique$num_unique,
    shannon = shannon,
    simpson = simpson,
    inv_simpson = inv_simpson
)

# Reset the row names to default
rownames(final_dataset_template) <- NULL

# Add coordinates from oba data to final dataset
add_coordinates <- left_join(final_dataset_template,
                             select(oba_data_subset,
                                    locality,
                                    decimalLatitude,
                                    decimalLongitude),
                             by = join_by(place_name == locality)) %>%
                   group_by(place_name, num_unique, shannon, simpson, inv_simpson) %>%
                   summarize(latitude = list(decimalLatitude),
                             longitude = list(decimalLongitude),
                             .groups = "drop")

# Join the dataset with coordinates to the land use dataset
final_dataset <- left_join(add_coordinates,
                           land_use,
                           by = join_by(place_name))

# Write the dataset to the cleaned data folder as an RData file
saveRDS(final_dataset, file = here("data", "cleaned_data", "oba_land_use.rds"))
```