---
title: "Effects of Land Use in Census Places on Pollinator Diversity"
author: "Hunter Everton"
output: 
  pdf_document:
    latex_engine: xelatex
    keep_tex: false
mainfont: Times New Roman
bibliography: ../bibliography/references.bib
csl: ../bibliography/apa-6th-edition.csl
urlcolor: blue
---

# Introduction

The goal of this project was to assess the effects of human land use in Oregon's cities and Census designated places on local pollinator diversity.
The effects of land use on pollinator diversity has been widely studied across different contexts around the world. In a large-scale review of 
studies researchers have found positive correlations between land use intensity and reductions in alpha-diversity [@tsang_land_2025]. Another 
review also saw a positive correlation between land use intensity and overall reductions in pollinator diversity, though their results
found little correlation between land use type (i.e. land used for residential purposes vs. for industrial purposes) and pollinator diversity 
[@millard_global_2021].

While the effects of land use on bee diversity are well documented, most pollinator studies measure landscape context using ecological or remotely
sensed units, such as fixed-radius buffers, land-cover categories, or habitat suitability indices. Parallel work in urban ecology has shown that
neighborhood socioeconomic conditions can influence pollinator habitat through gardening practices, greenspace management, and vegetation structure,
but these studies are largely limited to individual metropolitan areas. Importantly, U.S. Census places—key units for describing settlement patterns,
governance, and human activity—have not been systematically used to analyze pollinator biodiversity, despite their relevance for conservation
implementation and planning. No statewide studies have integrated high-resolution pollinator surveys (e.g., the Oregon Bee Atlas) with census-place
boundaries and land-use characteristics to evaluate how human-defined spaces structure biological diversity. This leaves a significant knowledge gap
concerning how pollinator species richness is distributed across towns and cities and how those patterns relate to land use in ways that are meaningful
for policy and community conservation efforts.

Other studies have looked at the effect of land use on the biodiversity of other flora and fauna, with results being generally similar to those
that assessed pollinators only [@lisboa_assessing_2022; @newbold_global_2020; @semenchuk_relative_2022]. Others have found more mixed results
across different contexts, seeing more biodiversity loss in certain areas or in some animal species over others [@newbold_global_2020;
@jernakoff_effects_2023; @cabernard_biodiversity_2024]. These other studies show that generalizing the effects of land use on other animal species
to animal biodiversity as a whole could be a mistake and lead to different results than those seen among pollinators specifically.

# Data

- NHGIS Land Cover Summary (https://www.nhgis.org/environmental-summaries): Collected by the NLCD, summarized by the NHGIS into different census 
categories (counties, census tracts, places, county subdivisions).
- A text file of Oregon census places, their FIPS codes and their place names. The NHGIS data does not have place names only codes, therefore this
will be useful for adding place names to the land use data.
- OBA data: I plan to transform the OBA data through subsetting the data to combine with the land use dataset. Since the OBA data is individual
collection records across various census places, this data will have to be transformed such that the biodiversity of each census place in the land
use data is found in the OBA data.

Skills represented:
1. Regular expressions
2. Table manipulation
3. Data visualization
4. Joining datasets

## Cleaning the Data

Load the necessary libraries
```{r load_libraries, message = FALSE}
library(here)
library(tidyverse)
library(ggplot2)
library(sf)
library(terra)
library(stringr)
library(vegan)
library(modelsummary)
```

Load the datasets
```{r, message = FALSE}
oba_data <- read_csv(here("data", "raw_data", "OBA_2018-2024.csv"))
land_use <- read_csv(here("data", "raw_data", "nhgis_place2022_tl2022_nlcd2021.csv"))
or_places <- read_delim(here("data", "raw_data", "or_places.txt"), delim = "|")
```

### Cleaning Land Use Data

First I'll work on the land use data as that requires more cleaning. The land use dataset contains all cities and incorporated places in the U.S.,
so first let's subset the data to just those from Oregon. The dataset from NHGIS uses FIPS codes in its GISJOIN variable; using the table of FIPS
codes for Oregon, which can be found [here](https://www2.census.gov/geo/docs/reference/codes2020/place/st41_or_place2020.txt), I will subset the
GISJOIN variable to the Oregon FIPS codes.

```{r warning = FALSE}
# Find the row where the Oregon places start
or_start <- which(land_use$GISJOIN == "G41000275")
# Find the end of the Oregon places
or_end <- which(land_use$GISJOIN == "G41084600")

# Subset the land use data to the Oregon rows
land_use <- land_use[or_start : or_end, ]

# Check that the land use data and place names have the same number of places
nrow(land_use) == nrow(or_places)
```

Now that I have the land use dataset subset to just the Oregon places I want to add the place names to the land use dataset. I then want to join the
datasets on the GISJOIN variable.

```{r warning = FALSE}
# Add the GISJOIN column structure from the land use data to the Oregon places dataset
or_places$GISJOIN <- paste("G", or_places$STATEFP, "0", or_places$PLACEFP, sep = "")

# Make sure they are the same, this should be the same as the number of rows in either dataset
sum(or_places$GISJOIN == land_use$GISJOIN) == nrow(land_use)

# Join on GISJOIN variable and select columns needed
land_use <- land_use %>%
    left_join(or_places, by = "GISJOIN") %>%
    select("PLACENAME", "COUNTIES", starts_with("PROP"))

# Rename all columns
colnames(land_use) <- c("place_name",
                        "county",
                        "prop_open_water",
                        "prop_perennial_ice_snow",
                        "prop_developed_open_space",
                        "prop_developed_low_intensity",
                        "prop_developed_med_intensity",
                        "prop_developed_high_intensity",
                        "prop_barren_land",
                        "prop_deciduous",
                        "prop_evergreen",
                        "prop_mixed_forest",
                        "prop_shrub_scrub",
                        "prop_grassland_herbaceous",
                        "prop_pasture_hay",
                        "prop_cultivated_crops",
                        "prop_woody_wetlands",
                        "prop_emergent_herbaceous_wetlands")
```

The place names end in either "city" or "CDP", and the places across multiple counties are denoted with the separator "~~~". These values should
be removed to merge with the oba data.

```{r warning = FALSE}
land_use$place_name <- str_remove(land_use$place_name, "\\s(city|CDP)$")
land_use$county <- str_remove(land_use$county, "~~~.*")
land_use$county <- str_remove(land_use$county, "\\s(County)")
```

The land use data is now cleaned and ready for merging. Next I will clean the oba data.

### Cleaning OBA Data

```{r warning = FALSE}
# Subset to Oregon data
oba_data_oregon <- oba_data[oba_data$stateProvince == "OR", ]

# Subset to places in the land use dataset
oba_data_places <- oba_data_oregon[oba_data_oregon$locality %in% land_use$place_name, ]

# Select needed columns
oba_data_subset <- oba_data_places %>%
    select(locality,
           genus,
           specificEpithet,
           decimalLatitude,
           decimalLongitude) %>%
    filter(!is.na(specificEpithet) & !is.na(locality)) %>%
    mutate(species = paste(genus, specificEpithet)) %>%
    select(-genus, -specificEpithet)

# Sort the data by locality in alphabetical order
oba_data_subset <- arrange(oba_data_subset, locality)

# Count the species for each locality
# Pivot the data to have each species in a column
oba_data_counts <- oba_data_subset %>%
    count(locality, species) %>%
    pivot_wider(names_from = species,
                values_from = n)

# Turn the data into a matrix for calculating diversity
oba_matrix <- as.matrix(oba_data_counts[, -1])

rownames(oba_matrix) <- unique(oba_data_subset$locality)

dim(oba_matrix)
```

Now I have a matrix where each row represents a location in both the place and OBA datasets and each column represents a species of bee in the OBA dataset.
Next I will compute diversity metrics of this matrix and add start creating the final dataset.

### Creating the final dataset

```{r warning = FALSE}
# Get unique species for each locality
oba_data_unique <- oba_data_subset %>%
    group_by(locality) %>%
    summarize(num_unique = n_distinct(species))

# Change NA values to be 0
oba_matrix[is.na(oba_matrix)] <- 0

# Find the shannon diversity index for each locality
shannon <- diversity(oba_matrix, index = "shannon")

# Find the simpson diversity index
simpson <- diversity(oba_matrix, index = "simpson")
```

For creating the final dataset I want to add coordinates from the oba data so that I can create spatial plots from the data.

```{r warning = FALSE}
# Set the columns of the final dataset to the diversity indices gotten earlier
# and the place names from the oba dataset
final_dataset_template <- data.frame(
    place_name = unique(oba_data_subset$locality),
    num_unique = oba_data_unique$num_unique,
    shannon = shannon,
    simpson = simpson
)

# Reset the row names to default
rownames(final_dataset_template) <- NULL

# Add coordinates from oba data to final dataset
add_coordinates <- left_join(final_dataset_template,
                             select(oba_data_subset,
                                    locality,
                                    decimalLatitude,
                                    decimalLongitude),
                             by = join_by(place_name == locality)) %>%
                   group_by(place_name, num_unique, shannon, simpson) %>%
                   summarize(latitude = list(decimalLatitude),
                             longitude = list(decimalLongitude),
                             .groups = "drop")

# Join the dataset with coordinates to the land use dataset
oba_land_use <- left_join(add_coordinates,
                          land_use,
                          by = join_by(place_name))
```

# Analysis

We can see that the final combined dataset has columns for the longitude and latitude of each bee record, each diversity index and the proportion of 
each land type. For making analysis easier I will combine the undeveloped land types into one measure of undeveloped land and the developed land intensities
into one developed land proportion.

```{r warning = FALSE}
oba_land_use_mutated <- oba_land_use %>%
    mutate(prop_undeveloped = prop_open_water +
                              prop_perennial_ice_snow +
                              prop_barren_land +
                              prop_deciduous +
                              prop_evergreen +
                              prop_mixed_forest +
                              prop_shrub_scrub +
                              prop_grassland_herbaceous +
                              prop_pasture_hay +
                              prop_cultivated_crops +
                              prop_woody_wetlands,
           prop_developed = prop_developed_open_space +
                            prop_developed_low_intensity +
                            prop_developed_med_intensity +
                            prop_developed_high_intensity) %>%
    select(place_name, num_unique, shannon, simpson,
           prop_developed, prop_undeveloped,
           longitude, latitude)

head(oba_land_use_mutated)
```

## Summary Statistics

```{r warning = FALSE}
ggplot(oba_land_use_mutated) +
    geom_histogram(aes(x = shannon), color = "white", fill = "black")

ggplot(oba_land_use_mutated) +
    geom_histogram(aes(x = simpson), color = "white", fill = "black")

ggplot(oba_land_use_mutated) +
    geom_histogram(aes(x = prop_developed),
                   color = "white", fill = "black") +
    labs(x = "proportion developed land")
```

Looking at the distribution of the statistics in the final dataset we can see that the shannon index is roughly normally distributed, the simpson
index is right skewed, and the proportion of developed land is roughly uniformly distributed. From the summary statistics themselves we can't see any
clear relationship between the proportion of developed land and either the shannon or simpson indices. This seems to be due to the fact that the proportion
of developed land is almost uniform across census places.

## Linear Regression

```{r warning = FALSE}
m_shannon <- lm(shannon ~ prop_developed, data = oba_land_use_mutated)

m_simpson <- lm(simpson ~ prop_developed, data = oba_land_use_mutated)

ggplot(oba_land_use_mutated, aes(x = prop_developed, y = shannon)) +
    geom_point() +
    geom_smooth(method = "lm") +
    labs(title = "Shannon Diversity", x = "proportion developed land", y = "shannon diversity index") +
    theme_bw()

ggplot(oba_land_use_mutated, aes(x = prop_developed, y = simpson)) +
    geom_point() +
    geom_smooth(method = "lm") +
    labs(title = "Simpson Diversity", x = "proportion developed land", y = "simpson diversity index") +
    theme_bw()

modelsummary(list("shannon" = m_shannon,
                  "simpson" = m_simpson),
             statistic = "p = {p.value}",
             output = "gt",
             gof_map = c("nobs", "r.squared"))
```

The linear regression models show the lack of relationship between the variables that I had considered after looking at the distributions of the variables.
We can see that the p-value for the relationship between the proportion of developed land and the shannon index is 0.207, and the p-value for the simpson index
is 0.463. These suggest no statistical significance for either linear model, which goes against the significant results seen in other studies of biodiversity
and land use intensity [@millard_global_2021; @lisboa_assessing_2022; @newbold_global_2020; @semenchuk_relative_2022]. The reasons for this are not entirely
clear and some ideas will be considered.

# Conclusion

From the analysis we can see that there is no statistical significance between the proportion of developed land and both the shannon and simpson indices.
This result is different from those in other studies, which found significant negative correlations between land use intensity and biodiversity [@millard_global_2021].
I hypothesize that this difference is due to census places not encapsulating the total diversity of pollinator communities or land use intensity. This
other study used a categorical variable for land use intensity rather than a proportion of total land so the method I used for measuring land use intensity
may have been insufficient for finding a significant relationship. Another idea is that Oregon is significantly different from the locations analyzed
in the other studies. This would require the methods of those other studies to be used in the context of Oregon, and I believe different results will
be seen if higher resolution data, like raster or vector spatial data from the NLCD, is used instead of census data aggregating data from the NLCD.

# References